Clustering based on probability distributions with application on residential customers

### 2021-10-17 main.Rmd
distance metrics section

The data set solely contains readings from smart meters and no information about the consumers' specific physical, geographical, or behavioural attributes. As a result, no attempt is made to explain why consumption varies. Instead, this work investigates how much energy usage heterogeneity can be found in smart meter data and what some of the most common electricity use patterns are. It is worth noting that when studying these dynamics, a variety of objectives may be pursued. One objective could be to group consumers with similar shapes over all relevant cyclic granularities. In this scenario, the variation in customers within each group is in magnitude rather than shape, while the variation between groups is only in shape. Most clustering algorithms offer only daily energy profiles throughout the hours of the day, but we suggest a broader approach to the problem, aiming to group consumers with similar shapes across all significant cyclic granularities. Another purpose of clustering could be to group customers that have similar differences in patterns across all major cyclic granularities, capturing similar jumps across categories regardless of the overall shape. For example, in the first goal, similar shapes across hours of the day will be grouped together, resulting in customers with similar behaviour across all hours of the day, whereas in the second goal, any similar big-enough jumps across hours of the day will be clubbed together, regardless of which hour of the day it is. Both of these objectives may be useful in a practical context and, depending on the data set, may or may not propose the same customer classification.

The proposed methodology aim to leverage the intrinsic temporal data structure hidden in time series data. The foundation of our method is unsupervised clustering algorithms based exclusively on time-series features. First, we study the underlying distributions that may have resulted in different patterns across temporal granularities in order to identify a mechanism to classify them based on the similarity of those distributions. Depending on the goal of clustering, the distance metric for defining similarity would be different. These distance metrics could be fed into a clustering algorithm to break large data sets into subgroups that can then be analyzed separately. These clusters may be commonly associated with real-world data segmentation. However, since the data is unlabeled a priori, more information is required to corroborate this. This section presents the work flow of the methodology:

@@

@wang2020tsibble introduced the tidy "tsibble" data structure to support exploration and modeling of temporal data. A tsibble comprises an index, optional key(s), and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the index of a tsibble is days, then a linear granularity might be weeks, months or years. For each key variable, the raw smart meter data is a sequence that is indexed by time and comprises values of several measurement variables at each time point. This sequence, though, could be depicted in a variety of ways. A shuffling of the raw sequence could reflect the distribution of hourly consumption over a single day, while another could indicate consumption over a week or a year. These temporal deconstructions of a time period into units such as hour-of-day, work-day/weekend are called cyclic temporal granularities. All cyclic granularities can be expressed in terms of the index set and could be augmented with the initial tsibble structure (index, key, measurements). It is worthwhile to note that the data structure changes while transporting from linear to cyclic scale of time as multiple observations of the measured variable would correspond to each category of the cyclic granularities. In this paper, quantiles are chosen to characterize the distributions for each category of the cyclic granularity. So, each category of a cyclic granularity corresponds to a list of numbers which is essentially few chosen quantiles of the multiple observations.

@@

The existing work on clustering probability distributions assumes we have an iid sample $f_1(v),\dots,f_n(v)$, where $f_i(v)$ denotes the distribution from observation $i$ over some random variable $v = \{v_t: t = 0, 1, 2, \dots, T-1\}$ observed across $T$ time points. In our work, we are using $i$ as denoting a customer and the underlying variable as the electricity demand. So $f_i(v)$ is the distribution of household $i$ and $v$ is electricity demand. In this work, instead of considering the probability distributions of the linear time series, we assume that the measured variables across different categories of any cyclic granularity are from data generating processes.
Hence, we want to be able to cluster distributions of the form $f_{i,A,B \dots, {N_C}}(v)$, where $A, B$ represent the cyclic granularities under consideration such that $A = \{a_j: j=1, 2, \dots J\}$,  $B = \{b_k: k  = 1, 2, \dots K\}$ and so on. We consider individual each category of a cyclic granularity ($A$) or combination of categories for interaction of cyclic granularities (for e.g. $A*B$) to have a distribution. For example, let us consider we have two cyclic  granularities of interest, $A = {0, 1, 2, \dots, 23}$ representing hour-of-day and  $B = \{Mon, Tue, Wed, \dots, Sun\}$ representing day-of-week. 
each customer $i$ consist of a collection of probability distributions. In case individual granularities ($A$ or $B$) are considered there are  $J = 24$ distributions of the form $f_{i,j}(v)$ or $K = 7$ distributions of the form $f_{i,k}(v)$ for each customer $i$. In case of interaction,  $J*K=168$ distributions of the form $f_{i,j, k}(v)$ could be conceived for each customer $i$. As a result, a distance between collections of these univariate probability distributions is required. Depending on the objective of the problem, there could be many approaches to considering such distances. This paper considers two approaches, which are explained in the next segment.

@@

Considering each individual or combined categories of cyclic granularities as a data generating process lead to a collection of conditional distributions for each customer $i$. The (dis) similarity between each pair of observations should be obtained by combining the distances between these collections of conditional distributions such that the resulting metric is a distance metric, which could be fed into the clustering algorithm. Two types of distance metric is considered:

@@

**Inter-category distances**

@@

\noindent The total distance between two elements $x$ and $y$ is then defined as $$S^A_{x,y} = \sum_{j} D_{x,y}(A)$$ (sum of distances between each category $j$ of cyclic granularity A) or  $$S^{A*B}_{x,y} = \sum_j \sum_k D_{x,y}(A, B)$$ (sum of distances between each combination of categories $(j, k)$ of the harmony $(A, B)$. When combining distances from individual cyclic granularities $A$ and $B$, $$S^{A, B}_{x, y} = S^A_{x,y}/J + S^B_{x,y}/K$$ is used, which could also be shown to be a distance metric easily. This is shown for cyclic granularity $A$ and $B$, but could be practically extended to more granularities.

**Intra-category distances**

Compute weighted pairwise distances ($wpd$) (XXX reference) for all considered granularities for all objects. $wpd$ is designed to capture the maximum variation in the measured variable explained by an individual cyclic granularity or their interaction and is estimated by the maximum pairwise distances between consecutive between consecutive categories normalized by appropriate parameters. A higher value of $wpd$ indicates that some interesting pattern is expected, whereas a lower value would indicate otherwise.

\noindent Distance between objects is then taken as the euclidean distances between them with the granularities being the variables and $wpd$ being the value under each variable. Since Euclidean distance is chosen, the observations with high values of features ($wpd$ values) will be clustered together. The same holds true for observations with low values of features. Thus this distance matrix would be useful to group customers that have similar significance of patterns across different granularities.

@@

<!-- grouping probability distributions across a harmony. This clustering algorithm is adopted to remove or appropriately adjust for auto correlation and unequal length in the data. The method could be further extended by clustering probability distributions conditional on one or more cyclic granularities. The following are some of the advantages of our proposed method. -->

>>>>

<!-- The data set solely contains readings from smart meters and no information about the consumers' specific physical, geographical, or behavioural attributes. As a result, no attempt is made to explain why consumption varies. Instead, this work investigates how much energy usage heterogeneity can be found in smart meter data and what some of the most common electricity use patterns are. -->

The proposed methodology aim to leverage the intrinsic temporal data structure hidden in time series data. The foundation of our method is unsupervised clustering algorithms based exclusively on time-series features. First, we study the underlying distributions that may have resulted in different patterns across temporal granularities in order to identify a mechanism to classify them based on the similarity of those distributions. It is worth noting that when studying these similarities, a variety of objectives may be pursued. One objective could be to group consumers with similar shapes over all relevant cyclic granularities. In this scenario, the variation in customers within each group is in magnitude rather than shape, while the variation between groups is only in shape. Most clustering algorithms offer only daily energy profiles throughout the hours of the day, but we suggest a broader approach to the problem, aiming to group consumers with similar shapes across all significant cyclic granularities. Another purpose of clustering could be to group customers that have similar differences in patterns across all major cyclic granularities, capturing similar jumps across categories regardless of the overall shape. For example, in the first goal, similar shapes across hours of the day will be grouped together, resulting in customers with similar behaviour across all hours of the day, whereas in the second goal, any similar big-enough jumps across hours of the day will be clubbed together, regardless of which hour of the day it is. Both of these objectives may be useful in a practical context and, depending on the data set, may or may not propose the same customer classification.

Depending on the goal of clustering, the distance metric for defining similarity would be different. These distance metrics could be fed into a clustering algorithm to break large data sets into subgroups that can then be analyzed separately. These clusters may be commonly associated with real-world data segmentation. However, since the data is unlabeled a priori, more information is required to corroborate this. This section presents the work flow of the methodology:

@@

@wang2020tsibble introduced the tidy "tsibble" data structure to support exploration and modeling of temporal data comprising of an index, optional key(s), and measured variables. <!-- An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the index of a tsibble is days, then a linear granularity might be weeks, months or years.--> For each key variable, the raw smart meter data is a sequence that is indexed by time and comprises values of several measurement variables at each time point. This sequence, though, could be depicted in a variety of ways. A shuffling of the raw sequence could reflect the distribution of hourly consumption over a single day, while another could indicate consumption over a week or a year. These temporal deconstructions of a time period into units such as hour-of-day, work-day/weekend are called cyclic temporal granularities. All cyclic granularities can be expressed in terms of the index set and could be augmented with the initial tsibble structure (index, key, measurements). It is worthwhile to note that the data structure changes while transporting from linear to cyclic scale of time as multiple observations of the measured variable would correspond to each category of the cyclic granularities. In this paper, quantiles are chosen to characterize the distributions for each category of the cyclic granularity. So, each category of a cyclic granularity corresponds to a list of numbers which is essentially few chosen quantiles of the multiple observations.

@@

The existing work on clustering probability distributions assumes we have an iid sample $f_1(v),\dots,f_n(v)$, where $f_i(v)$ denotes the distribution from observation $i$ over some random variable $v = \{v_t: t = 0, 1, 2, \dots, T-1\}$ observed across $T$ time points. In our work, we are using $i$ as denoting a customer and the underlying variable as the electricity demand. So $f_i(v)$ is the distribution of household $i$ and $v$ is electricity demand. In this work, instead of considering the probability distributions of the linear time series, we assume that the measured variables across different categories of any cyclic granularity are from different data generating processes.
Hence, we want to be able to cluster distributions of the form $f_{i,A,B \dots, {N_C}}(v)$, where $A, B$ represent the cyclic granularities under consideration such that $A = \{a_j: j=1, 2, \dots J\}$,  $B = \{b_k: k  = 1, 2, \dots K\}$ and so on. We consider individual category of a cyclic granularity ($A$) or combination of categories for interaction of cyclic granularities (for e.g. $A*B$) to have a distribution. For example, let us consider we have two cyclic  granularities of interest, $A = {0, 1, 2, \dots, 23}$ representing hour-of-day and  $B = \{Mon, Tue, Wed, \dots, Sun\}$ representing day-of-week. 
Each customer $i$ consist of a collection of probability distributions. In case individual granularities ($A$ or $B$) are considered, there are  $J = 24$ distributions of the form $f_{i,j}(v)$ or $K = 7$ distributions of the form $f_{i,k}(v)$ for each customer $i$. In case of interaction,  $J*K=168$ distributions of the form $f_{i,j, k}(v)$ could be conceived for each customer $i$.

\noindent As a result, a distance between collections of these univariate probability distributions is required. Depending on the objective of the problem, there could be many approaches to considering such distances. This paper considers two approaches, which are explained in the next segment.

@@

Considering each individual or combined categories of cyclic granularities as a data generating process lead to a collection of conditional distributions for each customer $i$. The (dis) similarity between each pair of customers should be obtained by combining the distances between these collections of conditional distributions such that the resulting metric is a distance metric, which could be fed into the clustering algorithm. Two types of distance metric is considered:

@@

**JS-based distances**

@@

\noindent The total distance between two elements $x$ and $y$ is then defined as $$S^A_{x,y} = \sum_{j} D_{x,y}(A)$$ (sum of distances between each category $j$ of cyclic granularity A) or  $$S^{A*B}_{x,y} = \sum_j \sum_k D_{x,y}(A, B)$$ (sum of distances between each combination of categories $(j, k)$ of the harmony $(A, B)$. When combining distances from individual $L$ cyclic granularities $C_l$ with $n_l$ levels, $$S_{x, y} = \sum_lS^{C_l}_{x,y}/n_l$$ is used, which is also a distance metric being the sum of JS distances.

**wpd-based distances**

Compute weighted pairwise distances ($wpd$) for all considered granularities for all objects. $wpd$ is designed to capture the maximum variation in the measured variable explained by an individual cyclic granularity or their interaction and is estimated by the maximum pairwise distances between consecutive categories normalized by appropriate parameters. A higher value of $wpd$ indicates that some interesting pattern is expected, whereas a lower value would indicate otherwise.

\noindent Distance between elements is then taken as the euclidean distances between them with the granularities being the variables and $wpd$ being the value under each variable. Since Euclidean distance is chosen, the observations with high values of features ($wpd$ values) will be clustered together. The same holds true for observations with low values of features. Thus this distance matrix would be useful to group customers that have similar significance of patterns across different granularities.

@@

<!-- grouping probability distributions across a harmony. This clustering algorithm is adopted to remove or appropriately adjust for auto correlation and unequal length in the data. The method could be further extended by clustering probability distributions conditional on one or more cyclic granularities. The following are some of the advantages of our proposed method. --
